From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-08-01 WOLFHPC15/article/mscac.tex
Compilation + sec 3

The solution proposed in this paper contains a domain specific language to describe a multi-stencil program, \ie an overall numerical simulation, and a compiler which successively parses the front-end language, transforms the internal representation, and finally dumps the representation to a component-based back-end code.

The proposed DSL, named MSL (\emph{Multi-Stencil Language}), is an agnostic descriptive language for multi-stencil simulations. Agnostic means that the description of a numerical simulation does not depend on implementation choices as, for example, the type of mesh (structured, unstructured), or its associated interfaces (how to define a stencil shape, etc.). Descriptive, on the other hand, means that the terminal states of the language grammar are (string) identifiers. As a result, MSL is not made to handle the expression of numerical computations, but only the description of a multi-stencil simulation, in a coarse-grain fashion. %Numerical computations are encapsulated in specific components that users have to develop. 
In a MSL description, six main sections are defined, as the six tuples of the definition of a multi-stencil program of Equation~\ref{eq:msp}.

@@

As illustrated in Figure~\ref{fig:mscac}, the compiler, named MSCAC for \emph{Multi-Stencil Component Assembly Compiler}, is composed of three different phases. The first phase of the compiler is to \emph{parse} the MSL description. Second, the compiler analyzes and transforms the computation part of the description $\Gamma$, to an intermediate representation. This phase of compilation, called MSC, is responsible for the automatic parallelization of the overall numerical simulation. Finally, the parallel intermediate representation is dumped to an actual implementation of the parallel pattern of the simulation.

@@

%In this paper are presented the language MSL, the MSC compilation phase, and the dump compilation phase. 
The parallelization techniques handled by MSC are \emph{data parallelism}, where data are splitted among processors or cores while applying a single program everywhere, and \emph{task parallelism} where the program is transformed into a dependency graph of tasks. From a dependency graph, different strategies can be used to schedule tasks, in other words to control their execution in a valid order. The MSC compiler presented in this paper handles the transformation of the dependency graph to a static scheduling.

The final phase of compilation, the dump, uses the static schedule of the tasks, created from $\Gamma$, and additionnal information on $T$, $\mathcal{M}$, $\Delta$, $\mathcal{E}$ and $\mathcal{D}$, given by the \emph{parser}, to produce a back-end solution which represents the parallel pattern or skeleton of the overall simulation. We could have propose a dump to OpenMP~\cite{660313}, however this paper presents a first contribution to dump a DSL to a component-based runtime. Component models, where a component represents a clear and independent functionnality of the application, have first been proposed in software engineering and have proved many times good properties for code re-use, separation of concerns, maintainability and productivity of codes. Those properties, on the other hand, are recurrent stated problems of parallel applications.

>>>>

The solution proposed in this paper is based on a domain specific language to describe a multi-stencil program, \ie a complete numerical simulation, and a compiler which successively parses the front-end language, transforms the internal representation, and finally dumps the representation to a component-based back-end code.

The proposed DSL, named MSL (\emph{Multi-Stencil Language}), is an agnostic descriptive language for multi-stencil simulations. Agnostic means that the description of a numerical simulation does not depend on implementation choices as, for example, the type of mesh (structured, unstructured), or its associated interfaces (how to define a stencil shape, etc.). Descriptive, on the other hand, means that the terminal states of the language grammar are (string) identifiers. As a result, MSL is not made to handle the expression of numerical computations, but only the description of a multi-stencil simulation, in a coarse-grain fashion. %Numerical computations are encapsulated in specific components that users have to develop. 
A MSL description is made of six main sections that match the six tuples elements of the definition of a multi-stencil program of Equation~\ref{eq:msp}.
\todo[inline]{JB: dans une grammaire, les identifiants sont toujours des chaines de caracteres, je ne comprends pas}

@@

As illustrated in Figure~\ref{fig:mscac}, the compiler, named MSCAC for \emph{Multi-Stencil Component Assembly Compiler}, is composed of three different phases. The first phase \emph{parses} the MSL description into an intermediate representation (IR). The second phase called MSC transforms the computation part ($\Gamma$) of the IR in order to automatically introduce parallelism of the overall numerical simulation. Finally, the third phase dumps the IR to an actual implementation of the parallel pattern of the simulation.

@@

%In this paper are presented the language MSL, the MSC compilation phase, and the dump compilation phase. 
The parallelization techniques handled by MSC are \emph{data parallelism}, where data is split among processors or cores while applying the same computation on every piece of data, and \emph{task parallelism} where the program is transformed into a dependency graph of tasks. From a dependency graph, different strategies can be used to schedule tasks, in other words to control their execution in a valid order. The MSC compiler presented in this paper handles the transformation of the dependency graph to a static scheduling.

The final phase of compilation, the dump, uses the static schedule of the tasks, created from $\Gamma$, and additional informations on $T$, $\mathcal{M}$, $\Delta$, $\mathcal{E}$ and $\mathcal{D}$, given by the \emph{parser}, to produce a back-end solution which represents the parallel pattern or skeleton of the overall simulation. We could have proposed a dump to OpenMP~\cite{660313}, however this paper presents a first contribution to dump a DSL to a component-based runtime. Component models, where a component represents a clear and independent functionality of the application, have first been proposed in software engineering and have proved many times good properties for code re-use, separation of concerns, maintainability and productivity of codes. Those properties, on the other hand, are recurrent stated problems of parallel applications.
